Viscous Transport in Eroding Porous Media

### 2019-07-30 docs/transport/erosion_transport.tex
add modified figs in transport paper
small changes to tortuosity

\input{preambles.tex}
%\usepackage{showkeys}
\begin{document}

@@

In Section~\ref{s:transport}, we we analyze passive tracers suspended in
an eroded geometry.  Give such a geometry $\Omega$, a tracer $\ss(t)$
initialized at $\ss_0$ is governed by 
\begin{align}
  \label{eqn:trajectories}
  \frac{d\ss}{dt} = \uu(\ss), \quad \ss(0) = \ss_0.
\end{align}
To compute such a trajectory, we apply the standard fourth-order
Runge-Kutta method.  To evaluate $\uu(\ss)$, depending on the proximity
of $\ss$ to $\bd\Omega$, we either use the trapezoid rule or the
Barycentric quadrature rule as described in Section~\ref{s:fmm}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Anomalous Diffusion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The dynamics of fluid in porous media can be characterized in terms of
anomalous diffusion.  However, the anomalous diffusion rate depends on
the properties of the geometry.  In addition to the porosity, the
distribution of the grains affects the anomalous diffusion rates.  To
quantify this rate, we require tracer trajectories that are governed by
equation~\eqref{eqn:trajectories}.  Given a trajectory, we define
$\lambda_j(t)$ to be the distance travelled by the particle
\begin{align}
  \lambda_j(t) = \int_{0}^t \|\ss_j(\tau)\|\, d\tau, 
    \quad j=1,\ldots,N_p.
\end{align}
Having formed $N_p$ trajectories, the first and second ensemble moments
are
\begin{align}
  \langle \lambda \rangle (t) = 
    \frac{1}{N_p} \sum_{j=1}^{N_p} \lambda_j(t), \quad 
    \sigma_\lambda^2(t) = \frac{1}{N_p} 
    \left[\lambda_j(t) - \langle \lambda \rangle(t) \right]^2.
\end{align}
If the particle spreading scales with time as $\sigma_\lambda^2 \sim
t^\alpha$, with $\alpha < 1$, this indicates that the particle
trajectories are sub-diffusive.  In contrast, if $\sigma_\lambda^2 \sim
t^2$, the particle trajectories are undergoing a ballistic motion and
are not affected by the grains.

For early times when the geometry has not eroded, we expect the
trajectories to be sub-diffusive.  However, as the bodies erode and
straight preferential channels are formed, we expect the dynamics to
resemble a ballistic motion.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Tortuosity}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The tortuosity of a geometry quantifies the average amount of turning
made by a streamline.  Mathematically, it is defined as
\begin{align}
  T = \frac{1}{d}\frac{\displaystyle\int_{S}u_x(y)\lambda(y)\,dy}
  {\displaystyle\int_{S}u_x(y)\,dy},
  \label{eqn:tortuosity1}
\end{align}
where $S$ is an cross-section perpendicular to the $x$ axis with length
$d$, $\lambda(y)$ is the length of the streamline that is initialized
at the $y$-coordinate along $S$, and $u_x(y)$ is the $x$ component of
the velocity at the $y$-coordinate along $S$.  Following~\cite{}, the
trajectories are formed until they cross a second cross-section.  In our
work, we always initialize the trajectories at $x=-1$ and stop them when
they reach $x=1$.

Under the assumptions that the flow is incompressible and not reentrant,
the tortuosity can be equivalently measured as~\cite{dud-koz-mat2011}
\begin{align}
  T = \frac{\displaystyle\int_\Omega \|\uu(\yy)\|\, d\yy}
           {\displaystyle\int_\Omega u_x(\yy)\, d\yy}.
  \label{eqn:tortuosity2}
\end{align}
The flow is reentrant if all streamlines connect the inlet to the
outlet.  Recirculation zones are possible in viscous fluids, but they
only constitute a small fraction of the total porous media.  Therefore,
the tortuosity defined in equation~\eqref{eqn:tortuosity2} is an upper
bound for the definition in equation~\eqref{eqn:tortuosity1}.  We have
used formulas for a particular geometry, and the difference is
negligible.  Since equation~\eqref{eqn:tortuosity2} does not require the
additional work of computing particle trajectories, we use this
definition for the tortuosity.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Streamlines and Tortuosity}
\label{s:streamlines}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\todo[inline]{Dump of earlier section for now ... need to fix how this
is worked into the section}
Having developed numerically accurate methods to compute the velocity
$\uu$ at points close to the bodies, we are able to compute the
trajectories of particles passing through an eroded body.  To find the
trajectory governed by equation~\eqref{eqn:trajectories}, we use the
fourth-order explicit Runge-Kutta method with a fixed time step size.
Since our method is both high-order in time and space, we are able to
simulate dynamics very close to the eroding bodies (Figure~\ref{fig:}
\todo[inline]{reference a figure or two from the results section)}.
Resolving the flow in these small velocity regions is essential to
accurately describe anomalous diffusion and
mixing~\cite{leb-den-dav-bol-car-dec-bou2011}.

To calculate tortuosity $T$, we use the simple quadrature rule
\begin{align}
  T = \frac{1}{d}\frac{\displaystyle\sum_{i=1}^{N} 
    u_x(y_i) \lambda(y_i) \Delta y_i}
  {\displaystyle\sum_{i=1}^{N} u_x(y_i) \Delta y_i}, 
\end{align}
where $y_i$ is the point on $S$ and $\Delta y_i = y_{i+1}-y_{i}$.

>>>>

\input{preambles.tex}
\usepackage{showkeys}
\begin{document}

@@

In Section~\ref{s:transport}, we we analyze passive tracers suspended in
an eroded geometry.  Give such a geometry $\Omega$, a tracer $\ss(t)$
initialized at $\ss_0$ is governed by 
\begin{align}
  \label{eqn:trajectories}
  \frac{d\ss}{dt} = \uu(\ss), \quad \ss(0) = \ss_0.
\end{align}
To compute such a trajectory, we apply the standard explicit
fourth-order Runge-Kutta method.  To evaluate $\uu(\ss)$, depending on
the proximity of $\ss$ to $\bd\Omega$, we either use the trapezoid rule
or the Barycentric quadrature rule as described in Section~\ref{s:fmm}.

@@

Having developed numerically accurate methods to compute the velocity
$\uu$ at points close to the bodies, we are able to compute the
trajectories of particles passing through an eroded body.  Since the
trajectories are computed with high-order methods in both space and
time,  we are able to simulate dynamics very close to the eroding bodies
(see Figures~\ref{fig:Eroding20tracer} and~\ref{fig:Eroding100tort}.
Resolving the flow in these small velocity regions is essential to
accurately describe anomalous diffusion and
mixing~\cite{leb-den-dav-bol-car-dec-bou2011}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Anomalous Diffusion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The dynamics of fluid in porous media is often characterized in terms of
anomalous diffusion~\cite{} \todo[inline]{need references}, and the
anomalous diffusion rate depends on the properties of the geometry.  The
anomalous diffusion depends on the porosity~\cite{}, but the
distribution of the bodies also affects the anomalous diffusion rates.
Since erosion causes certain structure in porous media, calculating the
anomalous diffusion rates in these geometries is important.

One way to define the anomalous diffusion rate is by using tracer
trajectories that are governed by equation~\eqref{eqn:trajectories}.
Given a set of trajectories, we define $\lambda_j(t)$ to be the distance
travelled by the particle
\begin{align}
  \lambda_j(t) = \int_{0}^t \|\ss_j(\tau)\|\, d\tau, 
    \quad j=1,\ldots,N_p.
\end{align}
Having formed $N_p$ trajectories, the first and second ensemble moments
are
\begin{align}
  \langle \lambda \rangle (t) = 
    \frac{1}{N_p} \sum_{j=1}^{N_p} \lambda_j(t), \quad 
    \sigma_\lambda^2(t) = \frac{1}{N_p} 
    \left[\lambda_j(t) - \langle \lambda \rangle(t) \right]^2.
\end{align}
At early times, the particles have not explored much of the geometry,
and we expect a ballistic motion $\sigma_\lambda^2 \sim t^2$.  However,
as the particles pass the grains, their trajectories are altered, and we
expect that $\sigma_\lambda^2 \sim t^\alpha$, with $\alpha < 1$, which
would indicate that the flow is sub-diffusive.  With respect to erosion,
as the bodies erode, we expect that the channelization of the geometry
results in more trajectories undergoing a ballistic motion, and this
would indicate a larger anomalous diffusion rate.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Tortuosity}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The tortuosity of a geometry quantifies the average amount of turning
made by a streamline.  The choice of how to average is not universal,
and Duda et al.~\cite{dud-koz-mat2011} summarize several options for
averaging.  One particular choice, which we shall adopt, allows us to
compute the tortuosity in terms of trajectories, or by using an area
integral over the fluid geometry.  Another advantage of this choice is
that it is not affected by regions of recirculation.  Mathematically,
the definition of tortuosity that we use is
\begin{align}
  T = \frac{1}{d}\frac{\displaystyle\int_{S}u_x(y)\lambda(y)\,dy}
  {\displaystyle\int_{S}u_x(y)\,dy},
  \label{eqn:tortuosity1}
\end{align}
where $S$ is an cross-section perpendicular to the $x$ axis with length
$d$, $\lambda(y)$ is the length of the streamline that is initialized
at the $y$-coordinate along $S$, and $u_x(y)$ is the $x$ component of
the velocity at the $y$-coordinate along $S$.
Following~\cite{dud-koz-mat2011}, the trajectories are formed until they
cross a second cross-section that is parallel to the initial
cross-section.  In our work, we always initialize the trajectories at
$x=-1$ and stop them when they reach $x=1$.

Under the assumptions that the flow is incompressible and not reentrant,
the tortuosity can be equivalently measured as~\cite{dud-koz-mat2011}
\begin{align}
  T = \frac{\displaystyle\int_\Omega \|\uu(\yy)\|\, d\yy}
           {\displaystyle\int_\Omega u_x(\yy)\, d\yy}.
  \label{eqn:tortuosity2}
\end{align}
The flow is reentrant if all streamlines connect the inlet to the
outlet.  Recirculation zones are possible in viscous fluids, but they
only constitute a small fraction of the total porous media and tend to
have very small velocities.  Therefore, the tortuosity defined in
equation~\eqref{eqn:tortuosity2} is an upper bound for the definition in
equation~\eqref{eqn:tortuosity1}, but for the geometries we consider, we
will show that their difference is negligible.  Since
equation~\eqref{eqn:tortuosity2} does not require the additional work of
computing particle trajectories, we use this definition for the
tortuosity.

The first method to compute numerically compute the tortuosity uses $N_p$ trajectories
that are initially at $(1,y_i)$, $i=1,\ldots,N_p$, and stopped when
their $x$ coordinate is 1.  T

with originating at $x=-1$ and terminating at $x=1$, we use the approximation 
\begin{align}
  T = \frac{1}{d}\frac{\displaystyle\sum_{i=1}^{N_p} 
    u_x(y_i) \lambda(y_i) \Delta y_i}
  {\displaystyle\sum_{i=1}^{N_p} u_x(y_i) \Delta y_i}, 
\end{align}
where $y_i$ is the point on $S$ and $\Delta y_i = y_{i+1}-y_{i}$.
Similarly, to compute the tortuosity in terms of the Eulerian velocity
field, we use the approximation
\begin{align}
  T = \frac{\sum\limits_{i,j=1}^N \|\uu(\yy_{i,j})\|\Delta x_i \Delta y_j}
      {\sum\limits_{i,j=1}^N u(\yy_{i,j}) \Delta x_i\Delta y_j},
\end{align}
where $\Delta x_i = 1/N$

@@
